Fast and scalable Gaussian process modeling of stellar variability \vspace{-3\baselineskip}

### 2016-12-26 paper/ms.tex
fixing stupid rem bug
a few more words about kernel parameters

Gaussian Processes \citep[GPs;][]{Rasmussen:2006} are a class of stochastic
models parameterized by a mean function $\mu_\bvec{\theta}(\bvec{x})$ and a
covariance or ``kernel'' function $k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$
parameterized by the parameters $\bvec{\theta}$ and $\bvec{\alpha}$
respectively.
Under this model, the log-likelihood of observing a dataset
\begin{eqnarray}
\bvec{y} = \left(\begin{array}{ccccc}
    y_1\quad && \cdots\quad && y_N
\end{array}\right)^\T
\end{eqnarray}
at coordinates
\begin{eqnarray}
X = \left(\begin{array}{ccccc}
    \bvec{x}_1\quad && \cdots\quad && \bvec{x}_N
\end{array}\right)^\T
\end{eqnarray}
is
\begin{eqnarray}\eqlabel{gp-likelihood}
\ln{p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})} =
    -\frac{1}{2} {\bvec{r}_\bvec{\theta}}^\T\,{K_\bvec{\alpha}}^{-1}\,
        \bvec{r}_\bvec{\theta}
    -\frac{1}{2}\ln\det K_\bvec{\alpha}
    - \frac{N}{2} \ln{(2\pi)}
\end{eqnarray}
where
\begin{eqnarray}
    \bvec{r}_\bvec{\theta} = \left(\begin{array}{ccccc}
    y_1 - \mu_\bvec{\theta}(\bvec{x}_1)\quad && \cdots\quad &&
    y_N - \mu_\bvec{\theta}(\bvec{x}_N)
\end{array}\right)^\T
\end{eqnarray}
is the vector of residuals and the elements of the covariance matrix $K$ are
given by $[K_\bvec{\alpha}]_{ij} = k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$.
The maximum likelihood values for the parameters $\bvec{\theta}$ and
$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
$\bvec{\alpha}$ using a non-linear optimization routine \dfmtodo{examples and
CITE}.
Similarly, probabilistic constraints on $\bvec{\theta}$ and $\bvec{\alpha}$
can be obtained by multiplying the likelihood by a prior
$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov Chain Monte Carlo (MCMC;
\dfmtodo{CITE}) algorithm to sample from the posterior probability density.

@@

In the following Section, we present a method of improving this scaling that
we call the \genrp\ method.
The \genrp\ method requires using a specific model for the covariance
$k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$ and it has several limitations.
The method can only be applied to one-dimensional datasets.
When we say ``one-dimensional'' here, it means that the \emph{input
coordinates} $\bvec{x}_i$ are scalar, $\bvec{x}_i \equiv t_i$.\footnote{We are
using $t$ as the input coordinate because one-dimensional GPs are often
applied to time series data but this isn't a real restriction and the \genrp\
method can be applied to \emph{any} one-dimensional dataset.}
Furthermore, the covariance function for the \genrp\ method is ``stationary''.
This means that the function $k_\bvec{\alpha}(t_i,\,t_j)$ is only a function
of $\tau_{ij} \equiv |t_i - t_j|$.

@@

\subsection{Gaussian processes with Lorentzian power spectra}

Parameters:
\begin{eqnarray}\eqlabel{kernel}
k(\tau) &=& \sum_{j=1}^p \left[
    a_j\,\exp\left(-c_j\,\tau\right)\,\cos\left(-d_j\,\tau\right) +
    b_j\,\exp\left(-c_j\,\tau\right)\,\sin\left(-d_j\,\tau\right)
\right]
\end{eqnarray}
where $\tau = |\Delta t|$.
Equivalently, this can be written as
\begin{eqnarray}
k(\tau) &=& \sum_{j=1}^p \left[
    \frac{1}{2}(a_j + i\,b_j)\,\exp\left(-(c_j+i\,d_j)\,\tau\right) +
    \frac{1}{2}(a_j - i\,b_j)\,\exp\left(-(c_j-i\,d_j)\,\tau\right)
\right]\quad.
\end{eqnarray}

The power spectral density is computed by taking the Fourier transform of
\eq{kernel} to find
\begin{eqnarray}
S(\omega) = \sum_{j=1}^p \sqrt{\frac{2}{\pi}}
\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+({c_j}^2+{d_j}^2)^2}\quad.
\end{eqnarray}

For some $b_j=0$, the PSD for the $j$-th component becomes
\begin{eqnarray}
S_j(\omega) = \frac{1}{\sqrt{2\,\pi}}\,\frac{a_j}{c_j}\,\left[
    \frac{1}{1+\left(\frac{\omega-d_j}{c_j}\right)^2} +
    \frac{1}{1+\left(\frac{\omega+d_j}{c_j}\right)^2}
\right]
\end{eqnarray}
a pair of Lorentzian distributions with width $c_j$ centered on
$\omega = \pm d_j$.

\citep{Rybicki:1995, Ambikasaran:2015}

\subsection{Generalization to quasi-periodic covariance}

Our method.

@@

\appendix \label{appendixa}

>>>>

Gaussian Processes \citep[GPs;][]{Rasmussen:2006} are a class of stochastic
models parameterized by a mean function $\mu_\bvec{\theta}(\bvec{x})$ and a
covariance or ``kernel'' function $k_\bvec{\alpha}(\bvec{x}_i,\,\bvec{x}_j)$
parameterized by the parameters $\bvec{\theta}$ and $\bvec{\alpha}$
respectively.
Under this model, the log-likelihood of observing a dataset
\begin{eqnarray}
\bvec{y} = \left(\begin{array}{ccccc}
    y_1\quad && \cdots\quad && y_N
\end{array}\right)^\T
\end{eqnarray}
at coordinates
\begin{eqnarray}
X = \left(\begin{array}{ccccc}
    \bvec{x}_1\quad && \cdots\quad && \bvec{x}_N
\end{array}\right)^\T
\end{eqnarray}
is
\begin{eqnarray}\eqlabel{gp-likelihood}
\ln{p(\bvec{y}\,|\,{X,\,\bvec{\theta}},\,\bvec{\alpha})} =
    -\frac{1}{2} {\bvec{r}_\bvec{\theta}}^\T\,{K_\bvec{\alpha}}^{-1}\,
        \bvec{r}_\bvec{\theta}
    -\frac{1}{2}\ln\det K_\bvec{\alpha}
    - \frac{N}{2} \ln{(2\pi)}
\end{eqnarray}
where
\begin{eqnarray}
    \bvec{r}_\bvec{\theta} = \left(\begin{array}{ccccc}
    y_1 - \mu_\bvec{\theta}(\bvec{x}_1)\quad && \cdots\quad &&
    y_N - \mu_\bvec{\theta}(\bvec{x}_N)
\end{array}\right)^\T
\end{eqnarray}
is the vector of residuals and the elements of the covariance matrix $K$ are
given by $[K_\bvec{\alpha}]_{nm} = k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$.
The maximum likelihood values for the parameters $\bvec{\theta}$ and
$\bvec{\alpha}$ for a given dataset $(\bvec{y},\,X)$ can be found by
maximizing \eq{gp-likelihood} with respect to $\bvec{\theta}$ and
$\bvec{\alpha}$ using a non-linear optimization routine \dfmtodo{examples and
CITE}.
Similarly, probabilistic constraints on $\bvec{\theta}$ and $\bvec{\alpha}$
can be obtained by multiplying the likelihood by a prior
$p(\bvec{\theta},\,\bvec{\alpha})$ and using a Markov Chain Monte Carlo (MCMC;
\dfmtodo{CITE}) algorithm to sample from the posterior probability density.

@@

In the following Section, we present a method of improving this scaling that
we call the \genrp\ method.
The \genrp\ method requires using a specific model for the covariance
$k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ and it has several limitations.
The method can only be applied to one-dimensional datasets.
When we say ``one-dimensional'' here, it means that the \emph{input
coordinates} $\bvec{x}_n$ are scalar, $\bvec{x}_n \equiv t_n$.\footnote{We are
using $t$ as the input coordinate because one-dimensional GPs are often
applied to time series data but this isn't a real restriction and the \genrp\
method can be applied to \emph{any} one-dimensional dataset.}
Furthermore, the covariance function for the \genrp\ method is ``stationary''.
This means that the function $k_\bvec{\alpha}(t_n,\,t_m)$ is only a function
of $\tau_{nm} \equiv |t_n - t_m|$.

@@

To scale GP models to larger datasets, \citet{Rybicki:1995} presented a method
of computing \eq{gp-likelihood} in $\mathcal{O}(N)$ when the covariance
function is
\begin{eqnarray}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})
\end{eqnarray}
where $\{{\sigma_n}^2\}_{n=1}^N$ are the measurement uncertainties,
$\delta_{nm}$ is the Dirac delta, and $\bvec{\alpha} = (a,\,c)$.
The intuition behind this method is that, for the choice of $k_\bvec{\alpha}$,
the inverse of $K_\bvec{\alpha}$ is tridiagonal and can it can be computed
with a small number of operations for each data point.
Subsequently, \citet{Ambikasaran:2015} generalized this method to arbitrary
mixtures of exponentials
\begin{eqnarray}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J a_j\,\exp(-c_j\,\tau_{nm})\quad.
\end{eqnarray}
In this case, the inverse becomes more complicated but \eq{gp-likelihood} can
still be evaluated in $\mathcal{O}(J^2\,N)$ operations where $J$ is the number
of components in the mixture and $N$ is still the number of data points.

It turns out that this kernel function can be made even more general by
introducing complex parameters $a_j \to a_j+i\,b_j$ and $c_j \to c_j+i\,d_j$.
In this case, the covariance function becomes
\begin{eqnarray}\eqlabel{genrp-kernel-complex}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^N &&\left[
    \frac{1}{2}(a_j + i\,b_j)\,\exp\left(-(c_j+i\,d_j)\,\tau_{nm}\right)
        \right. \nonumber\\
    &&+\left.
    \frac{1}{2}(a_j - i\,b_j)\,\exp\left(-(c_j-i\,d_j)\,\tau_{nm}\right)
\right]
\end{eqnarray}
and, for this function, \eq{gp-likelihood} can still be solved with
$\mathcal{O}(J^2\,N)$ operations.
The details of this method and a few implementation considerations are
discussed in \app{implementation}.

\eq{genrp-kernel-complex} can also be expressed as
\begin{eqnarray}\eqlabel{genrp-kernel}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
        \right.\nonumber\\
    &&+ \left.
    b_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\sin\left(d_j\,\tau_{nm}\right)
\right] \quad.
\end{eqnarray}
The Fourier transform of this covariance function is the power spectral
density (PSD) of the model and it is given by
\begin{eqnarray}
S(\omega) = \sum_{j=1}^p \sqrt{\frac{2}{\pi}}
\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+({c_j}^2+{d_j}^2)^2}\quad.
\end{eqnarray}
The physical interpretation of this model isn't immediately obvious and we
will return to a more general discussion of the physical intuition in a moment
but we can start with one useful special case of this model.
If we set the imaginary amplitude $b_j$ for some component $j$ to zero, that
term of \eq{genrp-kernel} becomes
\begin{eqnarray}
k_j(\tau_{nm}) =
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
\end{eqnarray}
and the PSD for the this component is
\begin{eqnarray}
S_j(\omega) = \frac{1}{\sqrt{2\,\pi}}\,\frac{a_j}{c_j}\,\left[
    \frac{1}{1+\left(\frac{\omega-d_j}{c_j}\right)^2} +
    \frac{1}{1+\left(\frac{\omega+d_j}{c_j}\right)^2}
\right] \quad.
\end{eqnarray}
This PSD is the sum of two Lorentzian or Cauchy distributions with width
$c_j$ centered on $\omega = \pm d_j$.
This model can be interpreted intuitively as a quasiperiodic oscillator with
amplitude $A_j = a_j$, quality factor $Q_j = c_j$, and period $P_j = 2\,\pi /
d_j$.

@@

\appendix

\section{Implementation details}\sectlabel{implementation}
